Retrofitting Effect Handlers onto OCaml

### 2020-11-19 retro-concurrency.tex
Update results based on stack cache

Fibers are allocated on the heap using |malloc| and are |free|d when no longer
necessary. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20} as we observed in our
evaluation (\S\ref{sec:eval}) that directly using |mimalloc|~\cite{Leijen19}
outperforms a stack cache.

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-5mm}
\end{table}

@@

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 13.77$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), whereas effect handlers benefit from stack allocation of the frames.
Our concurrency monad (and other monadic concurrency libraries such as
Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to the
lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
memory load latency for the local NUMA node is 93.2ns as measured using the
Intel MLC tool~\cite{mlc}.

>>>>

Fibers are allocated on the heap using |malloc| and are |free|d when no longer
necessary. Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack
cache} of recently freed stacks in order to speed up allocation.

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-5mm}
\end{table}

@@

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), whereas effect handlers benefit from stack allocation of the frames.
Our concurrency monad (and other monadic concurrency libraries such as
Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to the
lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{23ns, 5ns, 11ns and 7ns}, respectively. For reference, the idle
memory load latency for the local NUMA node is 93.2ns as measured using the
Intel MLC tool~\cite{mlc}.
